---
title: "ESS 330 Final Project"
author:
  - name: "Yazeed Aljohani"
    affiliation: "Colorado State University"
    corresponding: true
  - name: "Josh Puyear"
    affiliation: "Colorado State University"
  - name: "Cade Vanek"
    affiliation: "Colorado State University"
---

# Introduction, Background, and Motivation

The COVID-19 pandemic drastically changed the way people lived, worked, and moved, especially during the early lockdown period in 2020. Globally, one of the most significant consequences of these behavioral shifts was a sharp and sudden drop in greenhouse gas (GHG) emissions, particularly carbon dioxide (CO₂) emissions from fossil fuels. According to Liu et al. (2020), global fossil CO₂ emissions declined by approximately 7.8% in the early months of the pandemic (\@liu2020covid). This decrease was driven largely by transportation sectors, as road transport emissions dropped by about 15.5% and aviation emissions declined by nearly 29%. These reductions were felt across continents, including in China, Europe, and the United States (\@forster2020current).

Transportation is the largest source of greenhouse gas emissions in the United States, accounting for nearly 29% of total emissions (\@epa2022). Within urban environments, where population density and economic activity are centered, the contribution of transportation to local carbon footprints is often even greater. As cities tussle with climate change and look for strategies to lower emissions, understanding how transportation emissions responded to reduced human activity during COVID-19 becomes an important starting point. Did cities with more public transportation infrastructure see greater drops in emissions? Were some cities more resilient to emissions rebounds in 2021–2022? These are the kinds of questions our project aims to explore.

This project focuses on analyzing transportation-related carbon emissions in five major U.S. counties (Los Angeles County, Cook County, Harris County, San Francisco County, and King County) from 2019 through 2022. These counties were selected based on their large populations, diverse transportation infrastructure, and the availability of emissions and mobility data. By comparing emissions across three key periods: pre-pandemic (2019), pandemic lockdown (2020), and post-lockdown recovery (2021–2022). We aim to better understand how transportation behavior affects regional emissions and how these patterns can inform future urban planning and climate policies.

Our objective is to quantify and compare changes in transportation-related CO₂ emissions in these five counties before, during, and after the COVID-19 lockdowns. We hypothesize that counties with stricter lockdowns and greater reliance on public transit, such as San Francisco County and King County, will show more substantial emissions reductions in 2020 and possibly more gradual rebounds in the following years. In contrast, counties with more car-dependent infrastructure, such as Harris County and Los Angeles County, may show smaller reductions and quicker rebounds. By analyzing how emissions changed in relation to mobility patterns, we hope to uncover how regional urban form, transportation infrastructure, and human behavior interact during times of major disruption.

To address this research question, we are using two primary datasets. First, we use the EPA’s National Emissions Inventory (NEI), which provides comprehensive emissions estimates, including on-road transportation sources. The NEI is compiled every three years, with the most recent versions available for 2017 and 2020. While annual data is not always available, these datasets still provide valuable estimates for major emissions sectors at the county level, which aligns directly with our selected regions. For our analysis, we are focusing on transportation-related CO₂ emissions specifically in Los Angeles County, Cook County, Harris County, San Francisco County, and King County.

Our methodology involves three main components. First, we will **visualize** emissions and mobility trends over time using R and `ggplot2`. This will help identify patterns, anomalies, or significant changes during the lockdown period and beyond. Second, we will use a one-way ANOVA to statistically test whether mean CO₂ emissions differ significantly between the three periods (2019, 2020, and 2022). This will allow us to test our hypothesis that emissions were significantly lower during the lockdown year.

Finally, we will explore correlations between mobility and emissions to determine whether reductions in transit and workplace movement are associated with emission changes. While we acknowledge the limitations of this approach, such as mismatched temporal resolution and varying data accuracy, it still offers a strong starting point for evaluating behavior-emissions relationships. If time permits, we may consider adding additional contextual variables, such as county population density, public transportation usage rates, or local lockdown policies, to refine our analysis.

This project is important for several reasons. First, it provides insight into how short-term behavioral changes impact transportation emissions, a sector that is difficult to decarbonize. Second, it allows us to evaluate which counties were most successful in reducing emissions and why, knowledge that can inform future policies aimed at promoting sustainable, low-carbon transportation. Third, it contributes to a growing body of research on the environmental consequences of the COVID-19 pandemic and the lessons we can learn for future resilience planning.

In conclusion, by combining emissions data from the EPA’s NEI with behavioral data from Google’s mobility reports, this project explores how five major U.S. counties responded environmentally to the pandemic. Our analysis will quantify emission changes, investigate behavioral correlations, and highlight opportunities for sustainable urban development in a post-pandemic world.

# Data Exploration

# 

```{r}
library(tidyverse)
library(sf)
library(tigris)
library(ggplot2)
library(patchwork)  
library(maps)
library(usdata)
library(ggpubr)

options(tigris_use_cache = TRUE)

```

The first dataset is an EPA summary of various pollutants organized geographically. For our purposes, we will be only analyzing the CO2 measurements. The second set is google mobility data taken during the COVID-19 pandemic, measuring how different counties increased or decreased transportation frequency and usage.

cleaning:
Mobility data are only from 2022, so it would be hard to compare against timeseries emissions data.
```{r}
# This dataset is also only for one year and gets a summary of CO2 emissions, primary through quaternary sources of emissions, other data

co2_data <- read_csv("data/co2_emissions.csv") %>%
  mutate(
   FIPS = str_pad(FIPS, 5, pad = "0"),
  Emissions_Tons = `Emissions (Tons)`
 )

# Mobility data are only from 2022, so it would be hard to compare against timeseries emissions data.
mobility_data <- read_csv("data/mobility_data.csv") %>%
  mutate(
    FIPS = str_pad(census_fips_code, 5, pad = "0")
  ) %>%
  group_by(FIPS) %>%
  summarize(
   avg_residential = mean(residential_percent_change_from_baseline, na.rm = TRUE)
  )
```

Joining the previously considered data by fips
```{r}
joined_data <- co2_data %>%
 inner_join(mobility_data, by = "FIPS")
  

us_counties <- counties(cb = TRUE, class = "sf") %>%
 mutate(FIPS = paste0(STATEFP, COUNTYFP))

map_data <- us_counties %>%
left_join(joined_data, by = "FIPS")
```

Plotting the data with a map (this takes a lot of memory)
```{r}
# Define US bounding box (continental US)
#this generated a file that was 1.4 GB, far to large to be practical to run

# map_co2 <- ggplot(map_data) +
  # geom_sf(aes(fill = Emissions_Tons), color = NA) +
 # scale_fill_viridis_c(option = "plasma", trans = "log10", na.value = "grey90") +
  #  coord_sf(
  #  xlim = c(us_bbox["xmin"], us_bbox["xmax"]),  # Longitude bounds
   # ylim = c(us_bbox["ymin"], us_bbox["ymax"]),  # Latitude bounds
#expand = FALSE  # Remove padding
#  ) +
#  labs(title = "CO₂ Emissions (Tons)", fill = "Emissions") +
#  theme_minimal()

# Plot with zoom
# map_mobility <- ggplot(map_data) +
 # geom_sf(aes(fill = avg_residential), color = NA) +
 # scale_fill_viridis_c(option = "cividis", na.value = "grey90") +
 # coord_sf(
 #   xlim = c(us_bbox["xmin"], us_bbox["xmax"]),  # Longitude bounds
  #  ylim = c(us_bbox["ymin"], us_bbox["ymax"]),  # Latitude bounds
#    expand = FALSE  # Remove padding
#  ) +
#  labs(title = "Avg. Residential Mobility Change", fill = "% Change") +
#  theme_minimal()

# co2_mob <- map_co2 + map_mobility +
#  plot_layout(ncol = 2) +
#  plot_annotation(title = "County-Level CO₂ Emissions vs. Mobility Change")

```

t test(previously considered data):
This suggests that CO2 levels and mobility are highly interrelated.
```{r}
#This code is categorizing emissions into the categoroies of high and low. Why? Why is an unpaired t-test performed? Why don't we have an f-test to test for variance first?

joined_data <- joined_data %>%
 mutate(emissions_group = if_else(Emissions_Tons >  median(Emissions_Tons, na.rm = TRUE), "High", "Low")) %>% 
  mutate(log_tons = log(Emissions_Tons))
  

t_test_result <- t.test(avg_residential ~ emissions_group, data = joined_data)

t_test_result

```

Linear regression shows a relationship between CO2 emissions (undated) and mobility in 2022

```{r}
model <- lm(avg_residential ~ Emissions_Tons, data = joined_data)

summary(model)

print(model)

#from lecture 9, plts residuals vs leverage
plot(model, which = 5)

ggboxplot(joined_data$log_tons)
gghistogram(joined_data$log_tons)

gghistogram(joined_data$avg_residential)

library(ggplot2)

#create scatterplot with fitted regression line
ggplot(model, aes(x = Emissions_Tons, y = avg_residential)) + 
  geom_point() +
  stat_smooth(method = "lm")

#I was trying to find out why some of the average emissions values were negative to see if it made sense, but I couldn't determine maximum emissions because there was NaN in the columns.

joined_data %>% 
  arrange(avg_residential, na.rm = TRUE) %>% 
  head()

```



# EDA for Country Data

```{r, echo = TRUE}
library(readr)

country_co2 <- read_csv('data/owid-co2-data.csv')

country_info <- read_csv('data/owid-co2-codebook.csv')

```
# Country CO2 EDA
```{r, echo = TRUE}
library(visdat)
library(ggplot2)
library(ggpubr)
library(visdat)
library(dplyr)

names(country_co2)
# We're most concerned with the columns country, year, iso_code, population, gdp, co2, total_ghg, oil_co2_per_capita, co2_per_capita, share_global_co2, and energy_per_capita

View(country_info)
```

# Five most populous countries with population
```{r, echo = TRUE}
country_co2 %>% 
  arrange(-population) %>% 
  filter(!is.na(iso_code)) %>% 
  select(population, year, country, iso_code) %>% 
  filter(year == 2023) %>% 
  head()

```

# Methods Identified
```{r, echo = TRUE}
country_co2_75yr <- country_co2 %>% 
  select(country, year, iso_code, population, gdp, co2, total_ghg, oil_co2_per_capita, co2_per_capita, share_global_co2, energy_per_capita, ghg_excluding_lucf_per_capita) %>% 
  filter(year > 1950) %>% 
  filter(iso_code %in% c('IND', 'CHN', 'IDN', 'PAK', 'USA'))
  
vis_dat(country_co2_75yr)

#The combined population of the five most populous countries is 4.3 billion people.
#These countries are India, China, United States, Indonesia, and Pakistan
```
there is some missing data, but all of the co2 data are accounted for

```{r, echo = TRUE}
#make a combined plot of the top 5 most populous countries

# Check distribution of CO2 emitters across countries in the last 75 years
#narrow it down to the timeframe of 2015 to 2023 for context in the top 5 CO2 emitters

```

```{r, echo = TRUE}
#scatterplots- # make a combined polot of the top 5 CO2 emitters

```

```{r, echo = TRUE}
#rrange histograms and boxplots of data

```

```{r, echo = TRUE}
#plot a relationship between CO2 and mobility- we are looking at a classifying model if we go with modeling, although now that we've taken timeseries data a predictive model might be the way to go.
```

```{r, echo = TRUE}
summary(country_co2)

table(country_co2)

#measure normality of the distribution among all countries

#we know that there will be an upward trend in all countries across years

```


```{r, echo = TRUE}
#or.test(joined_table, variable and ghg emissions)

# qqplot for normal distribution

```

```{r, echo = TRUE}

#additional question- what were the most predictive variables of total CO2 emissions?

# Were the predictor variables before COVID the same as after COVID?

```

```{r, echo = TRUE}


```

Will be added soon
